Wealth dynamics in the presence of network structure and primitive cooperation

### 2019-07-14 manuscript/draftcssa19.tex
section 2 completed

The current work originated in our attempts to incorporate and tease out the effects of social structures in simple models of wealth dynamics in the presence of environmental stochasticity, and a simple form of resource pooling. These models were inspired by our search for analogs of the grid world agent-based models (ABMs) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. The original model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, allowed resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

@@

\label{giniKindaSucks}
% SD: In light of recent discussions, we may want to downplay Gini and up-play the robustness of agents to exogenous shocks.
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian noise driven linear dynamical systems, here constant growth rate; the structure is modeled by networks, here random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Although both the dynamical system and the network model are quite well understood, the precise interplay of structure and dynamics produces interesting emergent wealth distributions. To the best of our knowledge, this specific combination of network structure and social dynamics has not been discussed in the computational social science (CSS) or mathematical sociology literature, and we consider this the primary contribution of this work.

@@

Two categories of questions

~\cite{redner2001guide} ~\cite{rv}  ~\cite{wilhite} ~\cite{arthur}

@@

The model is realized in a discrete-time Julia simulation program, in which each agent is represented as a mutable struct (a dedicated area of memory whose contents change over time) that maintains its state. An agent's state includes its current amount of wealth and which proto (if any) it is a member of. Additionally, using the LightGraphs package~\cite{LightGraphs-2017}, each agent is associated with a node of a randomly generated ER graph. Protos are also represented as mutable structs, each of which contains a list of member agent IDs and a current wealth balance.

@@

% cite Bootstrap?
We chose Julia for its flexible type system, its speed of execution, its ease of programming, and the availability of useful packages, such as LightGraphs, Gadfly~\cite{Gadfly-2018} (plotting), and Bootstrap (confidence intervals).  Additionally, Julia easily allows the programmer to invoke code from R packages when necessary, as we did for Gini coefficient calculation with the R package DescTools~\cite{DescTools-2019}.

>>>>

The current work originated in our attempts to incorporate and tease out the effects of social structures in simple models of wealth dynamics in the presence of environmental stochasticity, and a simple form of resource pooling. Specifically, we are interested in studying how structure and dynamics interact to produce and sustain inequality, and influence robustness to scarcity shocks. We answer the former question using Gini coefficients; the latter question is answered using survival time type analysis. As we discuss in section 3, while Gini was found to be of limited value, survival time analysis produced more insights on both questions of inequality and robustness.

These models were inspired by our search for analogs of the grid world agent-based models (ABMs) of Friesen and Mudigonda~\cite{srimil} where foraging agents that pool their resources were shown to on-average outperform non-pooling agents. The original model used foraging to mix the population and create opportunities for interaction, and, when certain conditions were met, allowed resource sharing. Our model achieves agent interaction by postulating a static network structure which partially mixes the agents. Of primary interest is whether network effects alone can generate and sustain differences in economic outcomes of otherwise homogeneous agents.

@@

\label{giniKindaSucks}
% SD: In light of recent discussions, we may want to downplay Gini and up-play the robustness of agents to exogenous shocks.
The phenomenon under scrutiny is the emergence and evolution of economic inequality in societies with non-trivial social structures that enable economic interactions and cooperation mechanisms. The model we use to answer questions surrounding this phenomenon must possess a dynamic model of wealth accumulation, a model of social structure, and a suitably useful measure of inequality. The dynamics are modeled as Brownian noise driven linear dynamical systems, here constant growth rate; the structure is modeled by networks, here random graphs; the measures of inequality used are Gini coefficients, and survival time distributions of agents in response to lack of resources. Although both the dynamical system and the network model are quite well understood, the precise interplay of structure and dynamics produces interesting emergent wealth distributions and robustness to scarcity. To the best of our knowledge, this specific combination of network structure and social dynamics has not been discussed in the computational social science (CSS) or mathematical sociology literature, and we consider this the primary contribution of this work.

@@

The questions of interest to us, emergence and persistence of inequality and robustness to scarcity, in the presence of social structure and uncertain environmental conditions, map onto questions about the stochastic dynamical system (SDS). Inequality can be quantified using Gini coefficients of the particle ensemble's state space. Robustness to scarcity can be measured using survival time analysis for the particles to reach the absorbing boundary at the origin, by turning off the appropriate drift terms in the dynamics.

\subsection{Mathematical formulation}
The discussion of SDS closely follows ~\cite{redner2001guide}\footnote{The use of such analysis for studying wealth dynamics was presented by one of the authors at CSSA18 ~\cite{rv}. Unlike the finite interval dynamics used there, the dynamics here take place on a semi-infinite line.}

The SDS can be defined through a stochastic differential equation of the form

\begin{equation}\label{ddm}
dx(t) = vdt + \sqrt{2D} dw
\end{equation}

\noindent where $v$ is the wealth growth rate (the difference of the income and metabolic rate of the agent) and $D$ is the intensity of the Brownian process (white noise process) $w$. $x(t)$ is the state of the particle (wealth) at time $t$. The dynamics can be started at a any initial point $x_0 > 0$.

While equation (\ref{ddm}), in its discretized form, is the basis for simulations, other formulations are used~\cite{redner2001guide,rv}\footnote{These formulations make use measure theoretical probability to convert SDEs to partial differential equations known as Fokker-Planck equations.  The provide numerical and closed form estimates of probability density, survival time probabilities and other quantities of importance.}.

\noindent For a single particle, the probability of survival $S(t)$ and the expected probability of reaching the origin can be calculated and are functions of $x_0$, $v$ and $D$. For example, the expected probability of a particle reaching the origin ($\mathcal{E}(x_0)$), when starting at $x_0$ is given by

\begin{equation}\label{e}
\mathcal{E}(x_0) = \begin{cases}
e^{-v x_0/D} \text{, } & \text{ for } v > 0 \\
1 \text{, } & \text{ for } v \leq 0 \\
\end{cases}
\end{equation}

\noindent That is, there is always a non-zero probability of reaching the particle, even when there is a constant positive drift away from the origin; and, the return to origin is certain for negative or zero drift.  Similarly, survival probability, $\mathcal{S}(t)$ is given by

\begin{equation}\label{s}
\mathcal{S}(t) = \begin{cases}
1 - e^{-v x_0/D}\text{, } & \text{ } v >0 \\
\sqrt{\frac{4D}{\pi v^2 t} e^{-v^2 t/4 D}}\text{, } & \text{ } v \leq 0 \\
\end{cases}
\end{equation}

Since the particles are independent, they are independent dynamical systems; proto formation is a higher level construct imposed on this system and is a constraint on what states are considered viable and what are not. Consider two particles $x_1$ and  $x_2$ that have crossed the threshold for coalescence $x_{thresh}$. After the coalescence event, the state (wealth) of individual particles $x_1(t)$ and  $x_2(t)$ is not important; only the aggregate wealth of the coalescent (proto) is important. As long as $x_1(t) + x_2(t) >0$, both particles survive. Effectively, the proto is a two-dimensional SDS. Since the wealth of the proto ($p_{12}$)is additive, the aggregate wealth $p_{12} = x_1(t) + x_2(t) $. The dynamical variable satisfies an SDE\footnote{The result follows from the additivity properties of white noise.}

\begin{equation}\label{pddm}
dx(t) = 2vdt + \sqrt{2*2D} dw
\end{equation}

As one can see, only the coefficient of drift and diffusion in equations (\ref{ddm}) and (\ref{pddm}) are different. So, the corresponding expressions for $\mathcal{E}(x_0)$ and $\mathcal{S}(t)$ are suitably scaled. This has implications for both Gini coefficient calculations and survival analysis calculations.

For such ensemble of particles, the primary driver of difference in paths (life histories) and wealth is the Brownian noise intensity $D$; greater the $D$. greater the diversity (and hence Gini). However, Gini is a measure that is dependent on absolute magnitude. So,  if the drift ($v$) is large, then the variation generated by
$D$ gets washed out by $v$\footnote{Preliminary investigations suggest that for simple non-network ensembles, Gini either stays close to $0$ or $1$. We suspect that this is because of the constant wealth growth rate used in our models, Gini is a partially useful measure. This is unlike in macroeconomic models where exponential growth rate gives rise a appearance of non-trivial Gini coefficients}.  On the other hand, the mathematical form of expressions for $\mathcal{S}(t)$ suggest a clear and non-trivial dependence of the $D$ and clear separation between the population of particles that are isolates and the population of particles that are in a proto.

As the equations (\ref{e}) and (\ref{s}) show, all agents, irrespective of their starting intial condition and luck have a non-zero probability of dying. The paper that inspired this work ~\cite{srimil} emphasized studied inequality dynamics in the context of extreme scarcity in their foraging world. One can mimic this scarcity by simply turning of the income, leaving the agents to all inevitably die. The differential rate of death of the isolates and the non-isolates then become an important quantity.

In the simulation experiments that we discuss next discusses various stages and stopping conditions. All these stages and transition times between them make sense in terms of the underlying dynamics of the particle ensemble. In each of the stages, the effective dynamics are different from each other but the form of the equations remain the same.

To break the homogeneity of this particle ensemble, we propose to let the particles interact with only  a random subset of other populations. In other words, we embed the particles on a network (graph); particles interact (form proto) with only their neighbors. In this work, we focus on the simplest \textit{textbook} example of network: Erd\H{o}s-R\'{e}nyi network model~\cite{newman2018networks}. ER models has only one parameter $\lambda$, a Poisson parameter that determines the density of connections in the network, and in this setting, determines the number of particles available for a particle to coalesce and form protos. Commonsense reasoning suggests that a larger $\lambda$ parameter may help increase the average lifetimes of individual agents through by enabling greater likelihood of contact with other particles.

Just as equations (\ref{e}) and (\ref{s}) offers insight about the role of dynamics, the key feature of ER models is the the formation of giant component at $\lambda=1$, allowing a finite fraction of the population to be connected, have cliques of very high order, and other useful properties guarantee proto formation in large fraction of the population.

As its network properties are well-understood, wealth dynamics on such network allows us to understand more thoroughly, the interaction between dynamics and structure, and its role in global behavior. Still, a full analysis of this model remains to be completed and is part of a forthcoming work involving other more expressive network models. As we note subsequently, even this simple setting offers interesting insights,

@@

%The model is realized in a discrete-time Julia simulation program, in which each agent is represented as a mutable struct (a dedicated area of memory whose contents change over time) that maintains its state. An agent's state includes its current amount of wealth and which proto (if any) it is a member of. Additionally, using the LightGraphs package~\cite{LightGraphs-2017}, each agent is associated with a node of a randomly generated ER graph. Protos are also represented as mutable structs, each of which contains a list of member agent IDs and a current wealth balance.

@@

% cite Bootstrap?
%We chose Julia for its flexible type system, its speed of execution, its ease of programming, and the availability of useful packages, such as LightGraphs, Gadfly~\cite{Gadfly-2018} (plotting), and Bootstrap (confidence intervals).  Additionally, Julia easily allows the programmer to invoke code from R packages when necessary, as we did for Gini coefficient calculation with the R package DescTools~\cite{DescTools-2019}.
